# Causal Inference: Mediation and Instrumental Variables {#ch-7}

In previous chapters, we focused on establishing associations between variables. However, in health data science, we often want to understand the *causal* mechanisms driving these associations. This chapter introduces two advanced causal inference techniques: **Mediation Analysis** (to understand *how* an exposure affects an outcome) and **Instrumental Variables** (to estimate causal effects in the presence of unmeasured confounding).

::: {.yellowbox}
```{r, eval=FALSE}
devtools::install_github("BS1125/CMAverse")
```
:::

::: {.yellowbox}
```{r setup-ch7, message=FALSE, warning=FALSE}
# Load required libraries for this chapter
library(tidyverse)
library(gtsummary)
library(bda) 
library(AER)
library(CMAverse) 

# Note: We will simulate the data structures used in the examples below for rendering purposes.
```
:::

## Data Preparation for Causal Modeling

When investigating causal pathways, the timing of measurements is critical. An exposure must precede a mediator, which in turn must precede the outcome. Therefore, longitudinal data is typically required.

**Rationale & Code Explanation:** To prepare our data for regression and mediation, we often need to transform it from a "long" format (multiple rows per patient for different time points/waves) to a "wide" format (one row per patient, with variables denoting different time points, e.g., `bmi_1`, `bmi_5`). In the code below, we use `pivot_wider()` from the `tidyr` package to achieve this. We then select our variables of interest (age, sex, income, **\acr{BMI}**, and mental/physical wellbeing scores at different waves) and drop cases with missing data using `na.omit()`. 

::: {.yellowbox}
```{r data-prep-cleaning, eval=FALSE}
library(tidyverse)
library(gtsummary)

# 1. Raw data import and reshaping
dat <- read_tsv('../UKDA-8715-tab/tab/longitudinal_td.tab') %>% 
  select(pidp, indscus_lw_9, strata, psu,             
         wave, age_dv, sex_dv, fimngrs_dv, 
         bmi_dv, sf12pcs_dv, sf12mcs_dv) %>% 
  filter(wave %in% c(1, 5, 9)) %>%                    
  pivot_wider(id_cols=c(pidp, indscus_lw_9, strata, psu),                 
              names_from=wave, 
              values_from=age_dv:sf12mcs_dv) |> 
  select(-c(age_dv_5, age_dv_9, sex_dv_5, sex_dv_9, 
            fimngrs_dv_5, fimngrs_dv_9, bmi_dv_5, bmi_dv_9)) |> 
  na.omit()

# 2. Data formatting
dat <- dat %>% 
  transmute(
    age = age_dv_1, 
    sex = factor(sex_dv_1, labels=c('Male', 'Female')),  
    income = if_else(fimngrs_dv_1 > median(fimngrs_dv_1), 
                     'Higher income', 'Lower income'), 
    bmi = bmi_dv_1, 
    pw1 = sf12pcs_dv_1, mw1 = sf12mcs_dv_1, # Physical/Mental wellbeing Wave 1
    pw5 = sf12pcs_dv_5, mw5 = sf12mcs_dv_5, # Wave 5
    pw9 = sf12pcs_dv_9, mw9 = sf12mcs_dv_9  # Wave 9
  ) |> 
  filter(bmi >= 18.5)

# Optional: Save cleaned data
# write_csv(dat, '../Data/W7_USoc.csv')
```
:::

**Rationale:** Before running complex models, it is essential to inspect the descriptive statistics. The `tbl_summary()` function generates a clean summary table. We use the `statistic` argument to specify that we want the Mean and Standard Deviation (**\acr{SD}**) for all continuous variables, and the `digits` argument to round the results neatly.

::: {.yellowbox}
```{r descriptive-stats, eval=FALSE}
# Generate descriptive statistics table
tbl_summary(
  dat, 
  statistic = list(all_continuous() ~ '{mean} ({sd})'), 
  digits = list(all_continuous() ~ c(1, 1), 
                all_categorical() ~ c(0, 1))
)
```
:::

## Mediation Analysis

Mediation analysis attempts to unpack the **Total Effect** of an exposure on an outcome into two parts:
1. **Direct Effect:** The effect of the exposure directly on the outcome.
2. **Indirect Effect:** The effect of the exposure on the outcome that passes *through* a third variable (the mediator).

### The Base Model (Total Effect)

**Rationale:** Before looking for a mediator, we establish the baseline association. Here, we model mental wellbeing at Wave 9 (`mw9`) based on baseline **\acr{BMI}** (`bmi`), adjusting for demographic confounders and baseline wellbeing.

::: {.yellowbox}
```{r rq1-base-model, eval=FALSE}
# Base model predicting Wave 9 mental wellbeing
m0 <- lm(mw9 ~ bmi + age + sex + income + mw1 + pw1, data=dat)
tbl_regression(m0)
```
:::

### Regression Adjustment for Mediation

**Rationale:** The simplest way to check for mediation is the "difference method". We add our candidate mediator (Physical Wellbeing at Wave 5, `pw5`) into the regression model. If physical wellbeing mediates the relationship between **\acr{BMI}** and mental wellbeing, the coefficient for **\acr{BMI}** should drop noticeably in the new model compared to the base model. 

**Code Explanation:** We fit a second linear model (`m1`) that includes `pw5`. We then use `tbl_merge()` from the `gtsummary` package to put both models side-by-side for easy visual comparison. Minimal differences between the **\acr{BMI}** coefficients across the two models suggest minimal mediation.

::: {.yellowbox}
```{r rq2-regression-adj, eval=FALSE}
# Adjusted model including the mediator (pw5)
m1 <- lm(mw9 ~ bmi + age + sex + income + pw5 + mw1 + pw1, data=dat)

# Merge tables for side-by-side comparison
tbl_merge(
  list(tbl_regression(m0), tbl_regression(m1)), 
  tab_spanner = c('Not adjusted for PW', 'Adjusted for PW')
)
```
:::

### The Sobel Test

**Rationale:** The Sobel test is a traditional statistical test to determine if the indirect effect is significantly different from zero. 

**Code Explanation:** We use `mediation.test(mediator, exposure, outcome)` from the `bda` package. *Warning:* This test is rudimentary and does not adjust for any confounders. It is strictly not recommended for observational health data unless you are absolutely certain no confounding exists.

::: {.yellowbox}
```{r sobel-test, eval=FALSE}
library(bda)
# Sobel test (Unadjusted)
with(dat, mediation.test(pw5, bmi, mw9))
```
:::

### The Parametric G-Formula

Modern causal inference relies on more robust methods, such as the parametric G-formula, which can handle complex confounding scenarios (including interactions and non-linearities). In R, we use the `cmest()` function from the `CMAverse` package.

#### Without Exposure-Induced Confounders

**Rationale:** We define our causal pathways explicitly. The total effect (TE) is calculated by comparing an 'active' **\acr{BMI}** (`a = 26.69`, the sample mean) to a 'reference' **\acr{BMI}** (`astar = 25`). 

**Code Explanation:** * `basec`: Lists the baseline confounders.
* `mreg` and `yreg`: Specify the regression types for the mediator and outcome models (e.g., 'linear' or 'logistic').
* `mval`: Indicates a reference value for the mediator.
* `EMint=FALSE`: Specifies that we are not evaluating exposure-mediator interaction in this step.

::: {.yellowbox}
```{r gformula-no-confounder, eval=FALSE}
library(CMAverse)
# G-formula mediation analysis
mm1 <- cmest(
  data = dat, 
  model = 'gformula', 
  exposure = 'bmi', 
  mediator = 'pw5', 
  outcome = 'mw9', 
  basec = c('age', 'sex', 'income', 'mw1', 'pw1'), 
  a = 26.69, 
  astar = 25, 
  mreg = list('linear'), 
  mval = list(50), 
  yreg = 'linear', 
  EMint = FALSE
) 

summary(mm1)
```
:::

#### With Exposure-Induced Confounders

**Rationale:** In longitudinal data, an exposure (**\acr{BMI}** at Wave 1) might cause a change in a confounding variable (Mental Wellbeing at Wave 5), which then affects both the mediator and the final outcome. Traditional regression adjustment utterly fails in this scenario, but the G-formula can handle it.

**Code Explanation:** We add the `postc` argument to specify the intermediate (post-exposure) confounder (`mw5`). We also must tell R how to model this intermediate variable using the `postcreg` argument (here, as a 'linear' regression).

::: {.yellowbox}
```{r gformula-confounder, eval=FALSE}
# G-formula handling exposure-induced confounding (mw5)
mm2 <- cmest(
  data = dat, 
  model = 'gformula', 
  exposure = 'bmi', 
  mediator = 'pw5', 
  outcome = 'mw9', 
  basec = c('age', 'sex', 'income', 'mw1', 'pw1'), 
  postc = c('mw5'),       # The exposure-induced confounder
  a = 26.69, 
  astar = 25, 
  mreg = list('linear'), 
  mval = list(50), 
  postcreg = list('linear'), # Regression model for the confounder
  yreg = 'linear', 
  EMint = FALSE
) 

summary(mm2)
```
:::

*Interpretation Check:* If the Indirect Effect (IE) becomes significant and the mediation proportion jumps (e.g., to 34%) only after adjusting for `postc`, it implies that ignoring intermediate confounding severely underestimated the mediating role of physical wellbeing.

## Instrumental Variables (IV)

In observational data, or in clinical trials with non-compliance, unmeasured confounding can severely bias our effect estimates. Instrumental Variable (IV) analysis offers a workaround. An "instrument" is a variable that is strongly associated with the exposure, but has *no direct pathway* to the outcome except through the exposure itself. 

In a Randomized Controlled Trial (RCT) with poor adherence, the randomized `group` assignment is the perfect instrument: it dictates whether someone *should* exercise, but only the *actual attendance* impacts their stress levels.

### Intention-to-Treat (ITT) vs. Per-Protocol (PP)

**Rationale:** 1. **ITT Analysis:** We analyze patients based purely on the group they were randomized to, regardless of whether they actually attended the exercise sessions. This is statistically safe from confounding but usually underestimates the true biological effect of the intervention.
2. **Per Protocol (PP) Analysis:** We analyze patients based on their actual attendance. While this looks at the "true" intervention, it is heavily confounded (e.g., highly motivated people might attend more *and* have naturally lower stress).

::: {.yellowbox}
```{r itt-and-pp-analysis, eval=FALSE}
dat_iv <- read_csv('../Data/W7_IV.csv')

# 1. Intention-to-Treat (Conservative, unconfounded)
summary(lm(stress ~ group, data=dat_iv))

# 2. Per Protocol (Confounded by adherence behaviors)
summary(lm(stress ~ attendance, data=dat_iv))
```
:::

### IV Analysis using 2-Stage Least Squares

**Rationale:** IV analysis extracts the variation in `attendance` that is purely driven by the randomization `group` (the instrument), effectively purging the confounding noise from the model. 

**Code Explanation:** We use the `ivreg()` function from the `AER` package. The formula `stress ~ attendance | group` is read as: model the outcome (`stress`) against the exposure (`attendance`), using the variable after the pipe (`group`) as the instrument. 
We also test the strength of the instrument using `anova()`. An F-statistic > 10 generally indicates a strong instrument. Because `group` is a randomized allocation, it safely meets the exclusion criteria.

::: {.yellowbox}
```{r iv-analysis, eval=FALSE}
library(AER)

# Fit the Instrumental Variable regression
iv_model <- ivreg(stress ~ attendance | group, data=dat_iv)
summary(iv_model)

# Check instrument strength (F-statistic)
anova(glm(attendance ~ group, data=dat_iv), test='F')
```
:::